---
title: "project1"
Author: Jingyi Zuo
format: html
---

```{r store API}
library(httr)
library(jsonlite)
library(jsonlite)
# Store the API as an environment variable and reference it when access data.
  Sys.setenv(API_KEY = "b8e505a6fa8a9e78a5c7cbcf0bac66162486d10e")

```

# Write a function to GET() the URL, parse the JSON, and pull out a list of the variables with their values.

```{r Parsing Variable Values}
census_api <- function(year) {
# set url with input "year"
url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")

response <- GET((url),
      query = list(key=Sys.getenv("API_KEY")))
### obtain JSON text
content <- content(response, as = "text", encoding = "UTF-8")
###  parse the return
result <- fromJSON(content, flatten = FALSE)

### according to the returned content we parsed, we know the address of variables codebook is in c_variablesLink 
variable <- fromJSON(
content(GET(result$dataset$c_variablesLink,
      query = list(key=Sys.getenv("API_KEY"))), 
      as = "text",
      encoding = "UTF-8"),
            flatten = FALSE)
            
### set the variables we need in each year
var_state <- 
if (year %in% c(2021,2022)) "ST" 
else "STATE"

var_selected <- variable$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", var_state,"REGION", "DIVISION")]
return(var_selected)
}

### apply function in eachyear 
variable_2021 <- census_api(2021)
variable_2022 <- census_api(2022)
variable_2023 <- census_api(2023)
variable_2024 <- census_api(2024)

alldata <- list(variable_2021, variable_2022, variable_2023, variable_2024)
names(alldata) <- c("2021","2022","2023","2024")
saveRDS(alldata, file = "data/allyears.rds")
```


## Parsing API Data  

```{r parse API data and write a new function}
library(httr)
library(jsonlite)


url <- ("https://api.census.gov/data/2024/acs/acs1/pums")
response <- GET((url),
      query = list(key=Sys.getenv("API_KEY")))
### obtain JSON text
content <- content(response, as = "text", encoding = "UTF-8")
###  parse the return
result <- fromJSON(content, flatten = FALSE)
View(result$dataset) 





## helper 
transfer_tibble <- function(response) {
  data_content <- content(
    response,
    as = "text",
    encoding = "UTF-8"
  )

  result <- fromJSON(data_content)

#deal with colmmn_names
  column_names <- result[1, ]
  result <- result[-1, , drop = FALSE]
  colnames(result) <- column_names

  return(tibble::as_tibble(result))
}

##set up a query function, and the defalut value of year is 2024
query_census <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    variables = c("AGEP", "PWGTP", "SEX"),
    geography = "State",
    geography_values = NULL) {

legal_numeric_vars <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
)
legal_categorical_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")

if (!all(variables %in% c(legal_numeric_vars, legal_categorical_vars))) {
  stop("Invalid variable.")
}

numeric_vars <- variables[variables %in% legal_numeric_vars]
categorical_vars <- variables[variables %in% legal_categorical_vars]

if (!"PWGTP" %in% numeric_vars ||
    length(setdiff(numeric_vars, "PWGTP")) == 0) {
  stop("Include PWGTP and at least one other numeric variable.")
}

if (length(categorical_vars) == 0) {
  stop("Include at least one categorical variable.")
}  
    


