---
title: "project1"
Author: Jingyi Zuo
format: html
---

```{r store API}
library(httr)
library(jsonlite)
library(jsonlite)
# Store the API as an environment variable and reference it when access data.
  Sys.setenv(API_KEY = "b8e505a6fa8a9e78a5c7cbcf0bac66162486d10e")
  
library(lubridate)
  library(dplyr)
  library(quantreg) 

```

# Write a function to GET() the URL, parse the JSON, and pull out a list of the variables with their values.

```{r Parsing Variable Values}
census_api <- function(year) {
# set url with input "year"
url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")

response <- GET((url),
      query = list(key=Sys.getenv("API_KEY")))
### obtain JSON text
content <- content(response, as = "text", encoding = "UTF-8")
###  parse the return
result <- fromJSON(content, flatten = FALSE)

### according to the returned content we parsed, we know the address of variables codebook is in c_variablesLink 
variable <- fromJSON(
content(GET(result$dataset$c_variablesLink,
      query = list(key=Sys.getenv("API_KEY"))), 
      as = "text",
      encoding = "UTF-8"),
            flatten = FALSE)
            
### set the variables we need in each year
var_state <- 
if (year %in% c(2021,2022)) "ST" 
else "STATE"

var_selected <- variable$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", var_state,"REGION", "DIVISION")]
return(var_selected)
}

### apply function in eachyear 
variable_2021 <- census_api(2021)
variable_2022 <- census_api(2022)
variable_2023 <- census_api(2023)
variable_2024 <- census_api(2024)

alldata <- list(variable_2021, variable_2022, variable_2023, variable_2024)
names(alldata) <- c("2021","2022","2023","2024")
saveRDS(alldata, file = "data/allyears.rds")
```


## Parsing API Data  

```{r parse API data and write a new function}
library(httr)
library(jsonlite)


url <- ("https://api.census.gov/data/2024/acs/acs1/pums")
response <- GET((url),
                query = list(
                  get = "PWGTP,AGEP,SEX",
                  `for` = "state:12",
                  key = Sys.getenv("API_KEY")
))
### obtain JSON text
content <- content(response, as = "text", encoding = "UTF-8")
###  parse the return
result <- fromJSON(content, flatten = FALSE)
head(result)


## helper 
transfer_tibble <- function(response) {
  data_content <- content(
    response,
    as = "text",
    encoding = "UTF-8"
  )

  result <- fromJSON(data_content)

#deal with colmmn_names
  column_names <- result[1, ]
  result <- result[-1, , drop = FALSE]
  colnames(result) <- column_names

  return(tibble::as_tibble(result))
}

### turns value into times 
time_midpoint <- function(labels) {
clock_minutes <- function(clock) {
    clock <- gsub(".", "", clock, fixed = TRUE)
    parts <- strsplit(trimws(clock), "[: ]+")[[1]]

    hour <- as.numeric(parts[1]) %% 12
    if (parts[3] == "pm") hour <- hour + 12

    hour * 60 + as.numeric(parts[2])
  }

midpoint <- vapply(labels, function(label) {
    if (is.na(label) || !grepl(" to ", label, fixed = TRUE)) {
      return(NA_real_)
    }

    endpoints <- strsplit(label, " to ", fixed = TRUE)[[1]]
    start <- clock_minutes(endpoints[1])
    end <- clock_minutes(endpoints[2])

    if (end < start) end <- end + 1440

    ((start + end) / 2) %% 1440
  }, numeric(1))

return(
    lubridate::as_datetime(unname(midpoint) * 60, tz = "UTC")
  )
}

##set up a query function, and the default value of year is 2024
query_census <- function(
    year = 2024,
    variables = c("AGEP", "PWGTP", "SEX"),
    geography = "State",
    geography_values = NULL) {

legal_numeric_vars <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
)
legal_categorical_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")

if (!all(variables %in% c(legal_numeric_vars, legal_categorical_vars))) {
  stop("Invalid variable.")
}

numeric_vars <- variables[variables %in% legal_numeric_vars]
categorical_vars <- variables[variables %in% legal_categorical_vars]

if (!"PWGTP" %in% numeric_vars ||
    length(setdiff(numeric_vars, "PWGTP")) == 0) {
  stop("Include PWGTP and at least one other numeric variable.")
}

if (length(categorical_vars) == 0) {
  stop("Include at least one categorical variable.")
}  

## Check whether the input year is legal
if (length(year) != 1 || !year %in% 2021:2024) {
  stop("Year is illegal: Choose one year: 2021, 2022, 2023, or 2024.")
}

## Check geography level
if (length(geography) != 1 ||
    !geography %in% c("Region", "Division", "State")) {
  stop("Choose one geography level: Region, Division, or State.")
}


##Read  dictionary of the input year's
alldata <- readRDS("data/allyears.rds")
dictionary <- alldata[[as.character(year)]] 

##Choose the geography variable 
state_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"

## 

## Set default geography vafiables 
if (geography == "Region") {
  geography_var <- "REGION"
} else if (geography == "Division") {
  geography_var <- "DIVISION"
} else if (geography == "State") { 
  geography_var <- state_var
}

## Get legal geography codes from the dictionary
geography_labels <- unlist(
  dictionary[[geography_var]]$values$item
)

## add default values
if (is.null(geography_values)) {
  if (geography == "Region") {
    geography_values <- "1"
  } else if (geography == "Division") {
    geography_values <- "2"
  } else {
    geography_values <- "12"
  }
}

geography_values <- as.character(geography_values)

# Allow "all" in values, otherwise check the geography codes
if (length(geography_values) == 1 &&
    tolower(geography_values) == "all") {

  geography_codes <- "*"

} 
else {

  if (length(geography_values) == 0 ||
      !all(geography_values %in% names(geography_labels))) {
    stop("Invalid geography code.")
  }

  geography_codes <- paste(geography_values, collapse = ",")
}



### apply api request 
url <- paste0(
  "https://api.census.gov/data/", year, "/acs/acs1/pums"
)

response <- GET(
  url,
  query = list(
    get = paste(variables, collapse = ","),
    `for` = paste0(tolower(geography), ":", geography_codes),
    key = Sys.getenv("API_KEY")
  )
)


## apply hepler function 
data_tibble <- transfer_tibble(response)
for (variable in numeric_vars) {
  labels <- unlist(dictionary[[variable]]$values$item)
  values <- as.numeric(data_tibble[[variable]])

  if (variable %in% c("JWAP", "JWDP")) {
    time_labels <- labels[
      match(values, as.numeric(names(labels)))
    ]
    data_tibble[[variable]] <- time_midpoint(time_labels)
  } else {
    missing_codes <- names(labels)[grepl("^N/A", labels)]
    values[values %in% as.numeric(missing_codes)] <- NA_real_
    data_tibble[[variable]] <- values
  }
}

for (variable in categorical_vars) {
  labels <- unlist(dictionary[[variable]]$values$item)
  labels <- labels[order(as.numeric(names(labels)))]

  data_tibble[[variable]] <- factor(
    as.numeric(data_tibble[[variable]]),
    levels = as.numeric(names(labels)),
    labels = unname(labels)
  )
}

class(data_tibble) <- c("census", class(data_tibble))
return(data_tibble)
data_tibble <- transfer_tibble(response)
View(data_tibble)
}
```

```{r combine years}
query_census_years <- function(
    years = 2024,
    variables = c("AGEP", "PWGTP", "SEX"),
    geography = "State",
    geography_values = NULL) {

year_data <- lapply(years, function(year) {

### request data
census_data <- query_census(
      year = year,
      variables = variables,
      geography = geography,
      geography_values = geography_values
    )

## add year for variables 
census_data$year <- year
    return(census_data)
  })


return(dplyr::bind_rows(year_data))
}
combined_data <- query_census_years(
  years = c(2021, 2022, 2023, 2024)
)
combined_data$year <- factor(combined_data$year)

head(combined_data)
```


```{r numeric functions}
## we have returned a tibble; then we create a generic function for summaries 
summary.census <- function(
    tibble,
    numeric_vars = NULL,
    categorical_vars = NULL) {

#####check input, find out which variables should be suitable to be summarized. 
  ### if numeric variables are not specified explicitly, find out which variables in the tibble is numerical, and pass this into numeric_vars to prepare for summary. 
if (is.null(numeric_vars)) {
    numeric_vars <- names(tibble)[
      vapply(tibble, is.numeric, logical(1))
    ]
    numeric_vars <- setdiff(numeric_vars, c("PWGTP", "year"))
  }

# Ignore time variables
numeric_vars <- numeric_vars[
    !vapply(tibble[numeric_vars], function(x) {
      inherits(x, c("POSIXt", "difftime"))
    }, logical(1))
  ]

## if users didn't input any categorial variables, we take all the factor-type variables in the tibble to summarize. 
if (is.null(categorical_vars)) {
  categorical_vars <- names(tibble)[
    vapply(tibble, is.factor, logical(1))
  ]
}

# Treat year as categorical: if users input year as numeric, we take it into categorical (it has been transferred to factor)
if ("year" %in% numeric_vars) {
  numeric_vars <- setdiff(numeric_vars, "year")
  categorical_vars <- unique(c(categorical_vars, "year"))
}


## pass the parameters of which variable and its weight into different kinds of summariy function. 
numeric_summary <- lapply(numeric_vars, function(variable) {
    x <- tibble[[variable]]
    w <- tibble$PWGTP
## exclude all the observations with missing values
    keep <- !is.na(x) & !is.na(w)
    x <- x[keep]
    w <- w[keep]

    sample_mean <- sum(x * w) / sum(w)
    sample_sd <- sqrt(sum(x^2 * w) / sum(w) - sample_mean^2)

    c(mean = sample_mean, sd = sample_sd)
  })
  names(numeric_summary) <- numeric_vars

### categorical variables summarize   
  categorical_summary <- lapply(categorical_vars, function(variable) {
    x <- tibble[[variable]]
    w <- tibble$PWGTP
## exclude the observations with missing values 
    keep <- !is.na(x) & !is.na(w)

# apply function on each gorup 
    tapply(w[keep], x[keep], sum, default = 0)
  })
  names(categorical_summary) <- categorical_vars

  return(list(
    numeric = numeric_summary,
    categorical = categorical_summary
  ))
}
```

## plot 


``` {r plot}

plot.census <- function(x, ca_var, nu_var, ...) {

## check if users input one numeric and one categorical 
  if (length(ca_var) != 1 || length(nu_var) != 1) {
    stop("Specify only one categorical variable and only one numeric variable.")
  }

## check if users input really categorical, if not, output alarm 
  if (!is.factor(x[[ca_var]])) {
    stop("ca_var should be categorical.")
  }
## check if users input really numerical, if not, output alarm 
  if (!is.numeric(x[[nu_var]])){
    stop("nu_var should be numeric.")
  }

## return a plot: create a boxplot of the numeric variable at the levels of the categorical variable.
  
  
  return(
    ggplot2::ggplot(
      x,
      ggplot2::aes(
        x = factor(get(ca_var)),
        y = get(nu_var),
        weight = PWGTP
      )
    ) +
      ggplot2::geom_boxplot()
  )
}
```

## Test functions
Now we created four main functions: transfer_tibble(), query_census_years, summary.census()、plot.census(). We are going to test them.

``` {r investigation}
### Now we created 4 main functions for later using: query_census_years, summary.census()、plot.census() 

## We are going to test them.

##set up a query function, and the default value of year is 2024
query_1 <- query_census_years(
    years = c(2021,2022,2023,2024),
    variables = c("SCHL", "PWGTP", "SEX","AGEP","HHL"),
    geography = "State",
    geography_values = 12)


summary_query_1 <- summary.census(query_1)
summary_query_1

plot_school_age <- plot.census(query_1, ca_var = "HHL", nu_var = "AGEP")
  plot_school_age 

```
From the boxplot, we found that the age distribution is similar in household lanaguge groups, no clear pattern of the association between age and household language. 




```{r investigation about sex and education attainments }







